Making ice sheet models \\ scale properly

### 2023-01-24 talk/slides.tex
complete first draft I guess

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
an \alert{ice sheet model} is a \underline{map} which simulates an ice sheet in a climate
\end{definition} 
\item at least two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
\text{precipitation minus} \\
\text{melt \& runoff}
\end{pmatrix}$$

@@

\smallskip
\item {\small \alert{map: \, $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$} }
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}

@@

\medskip
\item conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] \emph{in} the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
    \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
\text{\alert{surface kinematic equation}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
\end{align*}

@@

\begin{itemize}
\item ice sheet evolution is a \alert{free-boundary} problem for conserved quantities
\item specifically, the surface kinematic equation
  $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
applies \emph{only} on the ice upper surface $\Gamma_s(t)$
\item in the remainder of the (fixed) domain $\Omega\subset \RR^2$, \alert{complementarity} holds:
  $$s=b \quad \text{and} \quad a \le 0$$

@@

\begin{itemize}
\item for this slide, forget complementarity and all boundary conditions
\item result is surface kinematic equation coupled to Stokes:
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
\nabla \cdot \bu &= 0
\end{align*}

\vspace{-2mm}
    \begin{itemize}
    \item[$\circ$] $s(t,x,y)$ and $\bu(t,x,y,z)$ both evolve in time
    \end{itemize}
\item<2-> only the first of these 5 equations has a time derivative
\item<3> the problem is a \alert{differential algebraic equation} (DAE) system:
\begin{align*}
\dot x &= f(x,y) \\
     0 &= g(x,y)
\end{align*}

    \begin{itemize}
    \item<3>[$\circ$] in infinite dimensions (PDAE?), and subject to complementarity
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{basic ice sheet model: current research and numerics}

@@

\medskip
\item<2> numerical modelers tend think of the Stokes problem separately from surface evolution
    \begin{itemize}
    \item[$\circ$] \emph{time-splitting} or \emph{explicit time-stepping} is taken for granted
    \end{itemize}
\item<2> glacier evolution is addressed with minimal awareness of complementarity
\end{itemize}
\end{frame}

@@

\section{time-stepping and stability}

@@

\begin{itemize}
\item thickness transport form helps expose evolution character
\item define
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

    \begin{itemize}
    \item[$\circ$] note: $s$ and $H$ are equivalent variables for modeling ice geometry
    \end{itemize}
\item<2-> the \alert{mass continuity equation} follows from the surface kinematical equation and incompressibility:
   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$

\smallskip
\item<3-> \emph{question:} is it really an advection?
\item<3->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H,$$
and ice is slow (DAE) so there are no characteristics
\end{itemize}
\end{frame}

@@

\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
\item the diffusion schema is only literal in the lubrication approximation
    \begin{itemize}
    \item[$\circ$] more on this momentarily
    \end{itemize}
\item the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences}, regardless of your advective schema preference
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item the basic ice sheet model (NCP coupled to Stokes) is \emph{not yet} in common use for ice sheet simulations
    \begin{itemize}
    \item[$\circ$] \alert{can we make it fast enough to use?}
    \end{itemize}

\smallskip
\item current-technology models use various shallow simplifications
    \begin{itemize}
    \item[$\circ$] usual justification: ``Stokes is too expensive at scale''
    \end{itemize}
\item the simplest of several shallow approximations is the ``lubrication'' approximation, the \alert{shallow ice approximation} (SIA)

@@

\begin{itemize}
\item in the shallow ice approximation, the surface motion contribution $\Phi(s) = - \bu|_s \cdot \bn_s$ has a formula:
{\small
\begin{gather*}
s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0 \\
{\color{FireBrick} \Phi(s) = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)}
\end{gather*}

\vspace{-2mm}
    \begin{itemize}
    \item[$\circ$] constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation
    \end{itemize}
}

\medskip
\item<2> $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
    \begin{itemize}
    \item<2>[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
    \item<2>[$\circ$] \emph{local} in surface and bed topography
    \item<2>[$\circ$] existence is known for this problem
         \begin{itemize}
         \item[$\vartriangleright$] Jouvet \& Bueler (2012); NCP is written as a \alert{variational inequality} weak form
         \end{itemize}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item the basic ice sheet model (NCP coupled to Stokes) problem actually has a \emph{non}-local surface velocity function $\Phi(s) = - \bu|_s \cdot \bn_s$
{\small
\begin{equation*}
s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0
\end{equation*}
}
\item \emph{figure:} the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the shallow ice approximation velocity responds only underneath the surface perturbation}
\end{itemize}

@@

\begin{frame}{time-stepping the mass continuity equation}

@@

\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{
\item \emph{caution}: for ice sheet models, $\bU(t,x,y)$ and $D(t,x,y)>0$ vary in space and time, from nonlinear coupling
\item nonetheless, let us recall some traditional numerical analysis of these equations \dots
}
\only<2>{
\item \alert{explicit} time stepping is common for \alert{advections}
\item for example, forward Euler using spacing $h$ and time step $\Delta t$:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{(\bU H)_{j+1/2}^\ell - (\bU H)_{j-1/2}^\ell}{h} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] conditionally stable, with maximum time step
        $$\Delta t \le \frac{h}{\max |\bU|} = O(h)$$
    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \end{itemize}
}
\only<3>{
\item \alert{explicit} time stepping for \alert{diffusions} is best avoided
\item for example, forward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell} + s_{j}^{\ell}) - D_{j-\frac{1}{2}} (s_{j}^{\ell} + s_{j-1}^{\ell})}{h^2} = a_j^\ell$$

@@

\begin{itemize}
\item current researchers mostly believe the advection schema
\item current-technology models mostly use explicit time stepping
    \begin{itemize}
    \item[$\circ$] embarrassing: \alert{the mathematical problem is a DAE}
    \item[$\circ$] time step size is determined by the coupled solution velocity $\bU$,
    \item[$\circ$] but there are performance and design consequences of hidden DAE \& diffusive character
    \end{itemize}

\medskip
\item<2> \alert{implicit time-stepping} is expected for DAE problems,
\item<2> but a sequence of difficult free-boundary problems must be solved at each time step: \alert{NCP coupled to Stokes at each time step}
    \begin{itemize}
    \item[$\circ$] not really ``equations'' any more because of $s\ge b$ inequality
    \end{itemize}
\end{itemize}
\end{frame}

\section{stress balance solver scaling}

@@

\begin{itemize}
\item assume an ice sheet model which is \emph{not} shallow
\item whether explicit or implicit, solves a Stokes problem at each step:
\begin{align*}
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item this problem is the \alert{stress balance} (conservation of momentum) which determines velocity $\bu$ and pressure $p$
\item how fast is the numerical solution process?
\end{itemize}
\end{frame}

\begin{frame}{PDE solver algorithmic scaling}

\begin{itemize}
\item as intro to scaling, consider the 3D Poisson equation:
    $$-\grad^2 u = f \text{ in $\Omega$}, \quad u=0 \text{ on $\partial\Omega$}$$
\item discretization generates a linear system $A\bu = \bb$
\item (data size $m$) = \alert{number of unknowns}
    \begin{itemize}
    \item[$\circ$] for low-order discretizations, $m$ = (number of nodes in the grid)
    \item[$\circ$] for cell size $h$, $m \sim h^{-3}$ in 3D
    \end{itemize}
\item solver choice determines asymptotic growth of computational expense (complexity)
\item \alert{algorithmic scaling} of flops as $m\to\infty$:
    \begin{itemize}
    \item[$\circ$] $O(m^3)$ for direct linear algebra, ignoring matrix structure
    \item[$\circ$] $\approx O(m^2)$ for sparsity-exploiting direct linear algebra
    \item[$\circ$] $O(m^1)$, \alert{optimal}, for \alert{multigrid} solvers (below)
    \end{itemize}
\end{itemize}

@@

\begin{frame}{scalable stress balance solvers for ice sheets}

\begin{itemize}
\item ice sheet stress balance solvers can be analyzed similarly
\item $m$ = (number of velocity and pressure unknowns)
    \begin{itemize}
    \item[$\circ$] indicate scaling by ``$O(m^{1+\alpha})$'', with $\alpha=0$ optimal
    \end{itemize}
\item \alert{near-optimal solvers} already exist: \hfill \emph{good news!}
    \begin{itemize}
    \item[$\circ$] $\alpha=0.08$ for Isaac et al.~(2015) Stokes solver
        \begin{itemize}
        \item[$\vartriangleright$] unstructured quadrilateral/tetrahedral mesh, $Q_k\times Q_{k-2}$ stable elements, Schur-preconditioned Newton-Krylov, column-oriented algebraic multigrid (AMG) preconditioner for $(\bu,\bu)$ block
        \end{itemize}

\begin{center}
\includegraphics[width=0.6\textwidth]{images/isaac-antarctica.png}
\end{center}
    \item[$\circ$] $\alpha=0.05$ for Tuminaro et al~(2016) 1st-order (shallow) AMG solver
    \item[$\circ$] similar for Brown et al~(2013) GMG solver
    \end{itemize}
\end{itemize}
\end{frame}

@@

@@

\begin{itemize}
\item for a time-stepping model our metric is \alert{flops per model year}
    \begin{itemize}
    \item[$\circ$] glaciologists want high-resolution simulations of ice sheets over 100k year ice age cycles
    \end{itemize}
\item assume map-plane domain $\Omega \subset \RR^2$ of width $L$
\item assume map-plane mesh cells have characteristic diameter $h$
\item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
\item the question addressed by upcoming performance table:

@@

\begin{frame}{preparation: observe and recall}

\begin{itemize}
\item ice sheets are thin layers, thus even non-shallow Stokes models often have $O(1)$ mesh points in the vertical direction
    \begin{itemize}
    \item[$\circ$] e.g.~Issac et al (2015) Stokes solver
    \item[$\circ$] \emph{message}: I am ignoring refinement in the vertical
    \end{itemize}

\medskip
\item scale relationships:
  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}} \hspace{20mm}$$

\medskip
\item mass-continuity equation time-stepping schemata:
{\footnotesize
\begin{align*}
\text{advective} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a & \Delta t &\le \frac{h}{U} \\
\text{diffusion} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a & \Delta t &\le \frac{h^2}{D}
\end{align*}
}

@@

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$
\end{tabular}

\bigskip\bigskip
\begin{itemize}
\item but what about implicit?
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item main point so far: \alert{explicit} time-stepping \alert{implies many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] yes, stress balance solver scaling ($\alpha$) is important
    \item[$\circ$] however, even optimality ($\alpha=0$) cannot fix performance
    \end{itemize}
\item switch to \alert{implicit time-stepping} for unconditional stability?
    \begin{itemize}
    \item[$\circ$] each step is a NCP-coupled-to-Stokes problem
        \begin{itemize}
        \item[$\vartriangleright$] a \alert{free-boundary problem}
        \item[$\vartriangleright$] table on next slide parameterizes cost of coupled implicit-step solves as $O(m^{1+\beta})$; optimal is $\beta=0$
        \end{itemize}
    \item[$\circ$] need $q$ model updates per year to integrate climate influences
    \end{itemize}

\bigskip
\item<2> \alert{existing} implicit time-stepping ice sheet models?
    \begin{itemize}
    \item[$\circ$] no fully-implicit NCP-coupled-to-Stokes models exist (?)
    \item[$\circ$] Bueler (2016) SIA NCP solver scales badly ($\beta=0.8$)
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$ \\
\\
implicit & \emph{either} & $\oo{\frac{q\, L^{2+2\beta}}{h^{\,2+2\beta}}} = \oo{q\, m^{1+\beta}}$
\end{tabular}
\end{frame}

\section{scalable approaches: candidates}

\begin{frame}{scalable approach 1: machine learning}

>>>>

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
an \alert{ice sheet model} is a map which simulates an ice sheet in a climate
\end{definition} 
\item at least two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
\text{precipitation minus} \\
\text{melt \& runoff}
\end{pmatrix}$$

@@

\smallskip
\item {\small \alert{map}: \, $\begin{pmatrix} \text{climate \&} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{\& velocity} \end{pmatrix}$ }
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}

@@

\medskip
\item conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] \emph{in} the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
    \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
\text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
\end{align*}

@@

\begin{itemize}
\item ice sheet evolution is a \alert{free-boundary} problem for conserved quantities
\item specifically, the surface kinematic equation (SKE)
  $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
applies \emph{only} on the ice upper surface $\Gamma_s(t)$
\item in the remainder of the (fixed) domain $\Omega\subset \RR^2$, \alert{complementarity} holds:
  $$s=b \quad \text{and} \quad a \le 0$$

@@

\begin{itemize}
\item for this slide, forget complementarity and boundary conditions
\item result is SKE coupled to Stokes:
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
\nabla \cdot \bu &= 0
\end{align*}

\item<2-> only the first of these 5 equations has a time derivative
    \begin{itemize}
    \item[$\circ$] because ice is slow and incompressible
    \end{itemize}
\item<3> the problem is a \alert{differential algebraic equation} (DAE), an extremely \alert{stiff}, system:
\begin{align*}
\dot x &= f(x,y) \\
     0 &= g(x,y)
\end{align*}

    \begin{itemize}
    \item<3>[$\circ$] but in $\infty$ dimensions (PDAE?) and subject to complementarity
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{basic ice sheet model: current research and thinking}

@@

\medskip
\item<2> numerical modelers tend to think of the Stokes problem separately from surface evolution
    \begin{itemize}
    \item[$\circ$] \emph{time-splitting} or \emph{explicit time-stepping} is taken for granted
    \end{itemize}
\item<2> ice sheet geometry evolution is addressed with minimal awareness of complementarity
\end{itemize}
\end{frame}

@@

\section{time-stepping}

@@

\begin{itemize}
\item thickness transport form helps expose evolution character
\item define:
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

    \begin{itemize}
    \item[$\circ$] note $s$ and $H$ are equivalent variables for modeling ice geometry
    \end{itemize}
\item<2-> the \alert{mass continuity equation} follows from the SKE and incompressibility:
   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$

\smallskip
\item<3-> \emph{question:} is this really an advection equation?
\item<3->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H,$$
and ice is slow (DAE) so there are no characteristics
\end{itemize}
\end{frame}

@@

\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item the diffusion schema is only literal in the lubrication approximation
    \begin{itemize}
    \item[$\circ$] more on this momentarily
    \end{itemize}
\item but the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences}
    \begin{itemize}
    \item[$\circ$] regardless of your advective schema preference
    \end{itemize}
\item in any case, both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item the basic ice sheet model (NCP coupled to Stokes) is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
    \begin{itemize}
    \item[$\circ$] \alert{can we make it fast enough to use?}
    \end{itemize}

\smallskip
\item current-technology models use various shallow simplifications
    \begin{itemize}
    \item[$\circ$] justification: ``Stokes is too expensive at scale''
    \end{itemize}
\item the simplest of several shallow approximations is the ``lubrication'' approximation, the \alert{shallow ice approximation} (SIA)

@@

\begin{itemize}
\item in the shallow ice approximation (SIA) version of the NCP,
{\small
  $$s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0$$
}
the surface motion contribution ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$ has a formula:
  $${\color{FireBrick} \Phi(s)} = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)$$

\vspace{-2mm}
    \begin{itemize}
    \item[$\circ$] constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation
    \end{itemize}

\medskip
\item<2> $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
    \begin{itemize}
    \item<2>[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
    \item<2>[$\circ$] \emph{local} in surface and bed topography
    \item<2>[$\circ$] existence is known for this NCP problem (Jouvet \& Bueler, 2012)
         \begin{itemize}
         \item[$\vartriangleright$] once written as a \alert{variational inequality} weak form
         \end{itemize}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item from now on, let's try to avoid shallowness approximations
\item then the basic ice sheet model (NCP coupled to Stokes) problem has a \alert{non-local} surface velocity function ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$
{\small
\begin{equation*}
s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0
\end{equation*}
}
\item \emph{figure:} the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the SIA velocity responds only underneath the surface perturbation}
\end{itemize}

@@

\begin{frame}{traditional PDE time-stepping}

@@

\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{
\item \emph{caution}: for ice sheet models, $\bU(t,x,y)$ and $D(t,x,y)>0$ vary in space and time, from nonlinear coupling
\item nonetheless, let us recall some traditional numerical analysis \dots
}
\only<2>{
\item \alert{explicit} time stepping is common for \alert{advections}
\item for example, forward Euler using spacing $h$ and time step $\Delta t$:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{(\bU H)_{j+1/2}^\ell - (\bU H)_{j-1/2}^\ell}{h} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] conditionally stable, with CFL maximum time step
        $$\Delta t \le \frac{h}{\max |\bU|} = O(h)$$
    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \end{itemize}
}
\only<3>{
\item \alert{explicit} time stepping for \alert{diffusions} is best avoided
\item for example, forward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell} + s_{j}^{\ell}) - D_{j-\frac{1}{2}} (s_{j}^{\ell} + s_{j-1}^{\ell})}{h^2} = a_j^\ell$$

@@

\begin{itemize}
\item most researchers believe the advection schema
\item current-technology models mostly use explicit time stepping
    \begin{itemize}
    \item[$\circ$] embarrassing: the mathematical problem is a DAE
    \item[$\circ$] time step size is determined by the coupled solution velocity $\bU$
         \begin{itemize}
         \item[$\vartriangleright$] CFL
         \end{itemize}
    \end{itemize}
\item however, the performance/usability/accuracy consequences of suppressed free-boundary/DAE/diffusive character are hard to sweep under the rug
\end{itemize}
\end{frame}

\begin{frame}{time-stepping in future ice sheet models}

\begin{itemize}
\item \alert{implicit time-stepping} should be used for DAE problems
\item a sequence of difficult free-boundary problems must be solved at each time step: \alert{NCP coupled to Stokes}
    \begin{itemize}
    \item[$\circ$] not really ``equations'', because of $s\ge b$ inequality
    \end{itemize}
\end{itemize}
\end{frame}

\section{stress-balance solver scaling}

@@

\begin{itemize}
\item whether explicit or implicit, a non-shallow model solves a Stokes problem at each step:
\begin{align*}
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item this is the \alert{stress balance} (conservation of momentum) problem which determines velocity $\bu$ and pressure $p$
\item how fast is the numerical solution process?
    \begin{itemize}
    \item[$\circ$] how do solution algorithms \alert{scale with increasing resolution}?
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{summary: PDE solver algorithmic scaling}

\begin{itemize}
\item consider the 2D Poisson equation:
    $$-\grad^2 u = f \text{ in $\Omega$}, \quad u=0 \text{ on $\partial\Omega$}$$
\item discretization generates a linear system $A\,\bu = \bb$
\item data size $m$ = (number of unknowns)
    \begin{itemize}
    \item[$\circ$] for low-order discretizations, $m$ = (number of nodes in the grid)
    \item[$\circ$] $m$ scales with mesh cell diameter $h$: \quad $m \sim h^{-2}$ in 2D
    \end{itemize}
\item \alert{complexity} or \alert{algorithmic scaling} of flops, as $m\to\infty$, depends on solver algorithm:
    \begin{itemize}
    \item[$\circ$] $O(m^3)$ for direct linear algebra, ignoring matrix structure
    \item[$\circ$] $\approx O(m^2)$ for sparsity-exploiting direct linear algebra
    \item[$\circ$] $O(m^1)$, \alert{optimal}, for \alert{multigrid} solvers (below)
    \end{itemize}
\end{itemize}

@@

\begin{frame}{ice sheet stress-balance solver complexity}

\begin{itemize}
\item $m$ = (number of velocity and pressure unknowns)
\item I'll parameterize scaling as $O(m^{1+\alpha})$, with $\alpha=0$ optimal
\item \alert{near-optimal solvers} already exist: \hfill $\leftarrow$ \emph{good news!}
    \begin{itemize}
    \item[$\circ$] $\alpha=0.08$ for Isaac et al.~(2015) Stokes solver
        \begin{itemize}
        \item[$\vartriangleright$] unstructured quadrilateral/tetrahedral mesh, $Q_k\times Q_{k-2}$ stable elements, Schur-preconditioned Newton-Krylov, ice-column-oriented algebraic multigrid (AMG) preconditioner for $(\bu,\bu)$ block
        \end{itemize}

\begin{center}
\includegraphics[width=0.6\textwidth]{images/isaac-antarctica.png}
\end{center}
    \item[$\circ$] $\alpha=0.05$ for Tuminaro et al~(2016) 1st-order (shallow) AMG solver
    \item[$\circ$] similar for Brown et al~(2013) 1st-order GMG solver
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{the analysis set-up: definitions and relationships}

\begin{itemize}
\item ice sheets are thin layers, thus ice sheet models often have $O(1)$ mesh points in the vertical direction
    \begin{itemize}
    \item[$\circ$] e.g.~Issac et al (2015) Stokes solver
    \item[$\circ$] \emph{message}: I am ignoring refinement in the vertical
    \end{itemize}

\medskip
\item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
\item relationships for map-plane domain $\Omega \subset \RR^2$ of width $L$ and cells of diameter $h$:
  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}} \hspace{20mm}$$

\item explicit time-stepping schemata:
{\footnotesize
\begin{align*}
\text{advective} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a & \Delta t &\le \frac{h}{U} \\
\text{diffusion} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a & \Delta t &\le \frac{h^2}{D}
\end{align*}
}

@@

\begin{itemize}
\item for a time-stepping model our metric is \alert{flops per model year}
    \begin{itemize}
    \item[$\circ$] glaciologists want high-resolution simulations of ice sheets over 100k year ice age cycles
    \end{itemize}

\bigskip
\item the question:

@@

@@

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$
\end{tabular}
\end{frame}

@@

\begin{itemize}
\item thus \alert{explicit} time-stepping \alert{implies many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] stress-balance solver scaling $\alpha$ is important, but even optimality ($\alpha=0$) cannot fix performance
    \end{itemize}
\item switch to \alert{implicit time-stepping} for unconditional stability?
    \begin{itemize}
    \item[$\circ$] need $q$ model updates per year to integrate climate influences
    \item[$\circ$] each step is a \alert{free-boundary} NCP-coupled-to-Stokes problem
    \item[$\circ$] parameterize cost of these solves as $O(m^{1+\beta})$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^{\,4}}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U \,L^{2+2\alpha}}{h^{\,3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{\,4+2\alpha}}} = \oo{\frac{D}{L^2} m^{\,2+\alpha}}$ \\
\\
implicit & \emph{either} & $\oo{\frac{q\, L^{2+2\beta}}{h^{\,2+2\beta}}} = \oo{q\, m^{1+\beta}}$
\end{tabular}
